Deep mutational scanning of hemagglutinin helps predict evolutionary fates of human H3N2 influenza variants

### 2018-03-31 paper/manuscript.tex

\begin{figure*}[ht]
\centering
\includegraphics[width=\textwidth]{figs/muteffect_by_maxfreq.pdf}
\caption{\label{fig:muteffect_maxfreq}
{\bf Experimental measurements are informative about the evolutionary fate of viral mutations.}
(A) Correlation between the effects of mutations as measured in our experiments and the maximum frequency reached by these mutations.
The plots show Spearman $\rho$ and an empirical $P$-value representing the proportion of 10,000 permutations for which permuted $\rho$ was greater than or equal to the observed $\rho$.
(B) The distribution of mutational effects partitioned by maximum mutation frequency.
The vertical black line shows the mean \comment{John: or is it median?} mutation effect for each category.
The analysis is performed separately for the pre- and post-Perth/2009 partitions of the tree (see Figure~\ref{suppfig:tree}).
}
\end{figure*}

@@

\begin{figure*}[ht]
\centering
\includegraphics[width=\textwidth]{figs/muteffect_by_maxfreq_WSN_1.pdf}
\caption{\label{fig:muteffect_maxfreq_WSN}
{\bf Experimental measurements on an H1 HA are not informative about the evolutionary fate of H3N2 mutations.}
This figure repeats the analysis of the H3N2 mutation frequencies in Figure~\ref{fig:muteffect_maxfreq}A, but uses the deep mutational scanning data for an H1 HA as measured in \cite{doud2016accurate}, i.e.\
Figure~\ref{suppfig:muteffect_maxfreq_WSN_supp} shows the histograms comparable to those in Figure~\ref{fig:muteffect_maxfreq}B.
\comment{John: Can you match any formatting changes that you make to Figure~\ref{fig:muteffect_maxfreq}A here so that they look consistent.}
}
\end{figure*}

@@

\subsection*{Inference of human H3N2 phylogenetic tree and calculation of maximum mutation frequencies}
To generate the tree shown in Figure~\ref{suppfig:tree}, we applied Nextstrain's augur pipeline~\cite{Hadfield224048} (\url{https://github.com/nextstrain/augur}; commit \texttt{006896d}) to publicly available H3N2 HA sequences from GISAID, sampling six viruses per month over the time interval of January 1, 1968 to April 1, 2017.
We aligned the resulting 2,139 HA sequences with MAFFT v7.310~\cite{katoh2013mafft} and constructed a maximum likelihood phylogeny from this alignment with RAxML 8.2.10~\cite{stamatakis2006raxml}.
Ancestral state reconstruction and branch length timing were performed with treetime~\cite{Sagulenko2018}.
We pruned seven strains with premature stop codons in their amino acid translations that represented likely sequencing errors, producing a final tree with 2,132 strains.
The phylogenetic tree is available as a JSON file on GitHub at \comment{repository will be made public upon acceptance of paper; if you are a reviewer who wants access please contact the editor}.
The tree was visualized using baltic (\url{https://github.com/blab/baltic}).

\comment{John: can you update this section to be accurate?}

\comment{John: can you add a section on calculation of max mutation frequencies?}

@@

>>>>

\begin{figure*}[ht]
\centering
\includegraphics[width=0.85\textwidth]{figs/muteffect_by_maxfreq.pdf}
\caption{\label{fig:muteffect_maxfreq}
{\bf Experimental measurements are informative about the evolutionary fate of viral mutations.}
(A) Correlation between the effects of mutations as measured in our experiments and the maximum frequency reached by these mutations.
The plots show Spearman $\rho$ and an empirical $P$-value representing the proportion of 10,000 permutations for which permuted $\rho$ was greater than or equal to the observed $\rho$.
(B) The distribution of mutational effects partitioned by maximum mutation frequency.
The vertical black line shows the mean \comment{John: or is it median?} mutation effect for each category.
The analysis is performed separately for the pre- and post-Perth/2009 partitions of the tree (see Figure~\ref{suppfig:tree}).
}
\end{figure*}

@@

\begin{figure*}[ht]
\centering
\includegraphics[width=0.85\textwidth]{figs/muteffect_by_maxfreq_WSN_1.pdf}
\caption{\label{fig:muteffect_maxfreq_WSN}
{\bf Experimental measurements on an H1 HA are not informative about the evolutionary fate of H3N2 mutations.}
This figure repeats the analysis of the H3N2 mutation frequencies in Figure~\ref{fig:muteffect_maxfreq}A, but uses the deep mutational scanning data for an H1 HA as measured in \cite{doud2016accurate}, i.e.\
Figure~\ref{suppfig:muteffect_maxfreq_WSN_supp} shows the histograms comparable to those in Figure~\ref{fig:muteffect_maxfreq}B.
\comment{John: Can you match any formatting changes that you make to Figure~\ref{fig:muteffect_maxfreq}A here so that they look consistent.}
}
\end{figure*}

@@

@@

\subsection*{Inference of human H3N2 phylogenetic tree and calculation of maximum mutation frequencies}
To generate the tree shown in Figure~\ref{suppfig:tree}, we applied Nextstrain's augur pipeline~\cite{Hadfield224048} (\url{https://github.com/nextstrain/augur}; commit \texttt{006896d}) to publicly available H3N2 HA sequences from GISAID \cite{shu2017gisaid}, sampling six viruses per month over the time interval of January 1, 1968 to April 1, 2017.
We aligned the resulting 2,139 HA sequences with MAFFT v7.310~\cite{katoh2013mafft} and constructed a maximum likelihood phylogeny from this alignment with RAxML 8.2.10~\cite{stamatakis2006raxml}.
Ancestral state reconstruction and branch length timing were performed with TreeTime~\cite{Sagulenko2018}.
We pruned seven strains with premature stop codons in their amino acid translations that represented likely sequencing errors, producing a final tree with 2,132 strains.
The phylogenetic tree is available as a JSON file on GitHub at \comment{repository will be made public upon acceptance of paper; if you are a reviewer who wants access please contact the editor}.
The tree was visualized using BALTIC (\url{https://github.com/blab/baltic}).

% notation for frequencies model
\newcommand{\dx}{\mathrm{d}x}						% change in x
\newcommand{\dy}{\mathrm{d}y}						% change in y
\newcommand{\dt}{\mathrm{d}t}						% change in t
\newcommand{\inertia}{\epsilon}			    % inertia parameter
\newcommand{\normal}{\mathcal{N}}				% normal distribution

The frequency trajectory of each individual mutation on the phylogeny is estimated following Nextstrain's augur pipeline and as first implemented in Nextflu \cite{neher2015nextflu}.
Herein, mutation frequency dynamics are modeled according to a Brownian motion diffusion process discretized to one-month intervals.
Relative to a simple Brownian motion, the expectation includes an `inertia' term $\inertia$ that adds velocity to the diffusion and the variance includes a term $x(1-x)$ to scale variance according to frequency following a Wright-Fisher population genetic process.
This results in the following diffusion process
\begin{equation}
x(t+\dt) = \normal\big( x(t) + \inertia \, \dx , \; \dt \, \sigma^2 \, x(t) \, (1-x(t)) \big),
\end{equation}
with `volatility' parameter $\sigma^2$.
The term $\dx$ is the increment in the previous timestep, so that $\dx = x(t) - x(t-\dt)$.
We used $\inertia = 0.7$ and $\sigma^2 = 0.05$ to maximize fit to empirical trajectory behavior.

We also include an Bernoulli observation model for mutation presence / absence among sampled viruses at timestep $t$.
This observation model follows
\begin{equation}
f(x,t) = \prod_{v\in V} x(t) \prod_{v\notin V} (1-x(t)),
\end{equation}
where $v\in V$ represents the set of viruses that have the mutation and $v\notin V$ represents the set of viruses that do not have the mutation.
Each frequency trajectory is estimated by simultaneously maximizing the likelihood of the process model and the likelihood of the observation model via adjusting frequency trajectory $\mathbf{x}=(x_1, \ldots, x_n)$.
